---
title: "Esm 206 Lab 5"
author: "Rachel Conway"
date: "10/29/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##### Lab 5 objectives

- Learn to parse dates with'lubridate'
- Get counts of obs w count()
- Uncount() rows
- One and two-sample t-test
- Create a heat map w geom_tile()

```{r, include= FALSE}

# Attach packages and read in data, clean up column names, change -99999 to NA

library(tidyverse)
library(here)
library(janitor)

lobster_abundance <- read_csv(here::here("data", "lobster_abundance.csv"), na= "-99999") %>% 
  janitor::clean_names()

```

Want each obs to have own row
Use tidyr::uncount() fxn to convert our lobster data from frequency format to case format

```{r}

lobster_tidy <- lobster_abundance %>%
  tidyr::uncount(lobster_count)

# data now truly in tidy format- uncount makes it easy to get from freq to case format, more ggplot friendly

```

#### Exploratory data visulization

We're only going to consider 'site' as our variable of interest

```{r}

ggplot(lobster_tidy, aes(x= site, y= size_mm)) +
  geom_jitter(aes(color= site),
              alpha= 0.5,
              width= 0.2)

# too many obs to see clear dist
# ask questions about groups-- seems have relatively simiar means-- tru to figure out where differences come from

# Now a histogram

ggplot(lobster_tidy, aes(x= size_mm)) +
  geom_histogram(aes(fill= site)) +
  facet_wrap(~site, scales= "free")

ggplot(lobster_tidy, aes(sample= size_mm)) +
  geom_qq()+
  facet_wrap(~site)

# assess distribution of points to determine if dist is normal, look at outliers
# by looking at hist and qq, observations look close to normal, only look at linear relationship w qq, slope means nothing
# even if samples did not look normally dist, might still be comfortable saying means are normally dist due to central limit theorem

```

#### Convert the 'date' column to class 'Date' 

using lubrudate package to convert to Date format, and then help us easily porse month and year

```{r}

lobster_date <- lobster_tidy %>%
  mutate(
    date_new= lubridate::mdy(date)
  )

# in console to check: class(lobster_date$date_new)

# now want to separate month and year so we can later group by those metrics
# use month and year fxns from lubridate

```

Now let's parse year and month using lubridate::month() and lubridate::year():

```{r}

lobster_parse_date <- lobster_date %>%
  mutate(
    obs_month= lubridate::month(date_new, label= TRUE),
    obs_year= lubridate::year(date_new)
  )

# in console to check: class(lobster_parse_date$obs_month), levels(lobster_parse_date$obs_month)

```

Questions about how many obs observed there-- different important groupings not based on freq counts

Now let's find counts of observed lobsters based on diff hierarchical groupings using the count fxn

First, let's count lobs by year and month
```{r}

lobster_ym <- lobster_parse_date %>%
  dplyr::count(obs_year, obs_month)

lobster_y <- lobster_parse_date %>%
  dplyr::count(obs_year)

lobster_site <- lobster_parse_date %>%
  dplyr::count(site)

# having data in tidy format allows us to create groupings based on what we find to be the most important parts of the data-- we as user of data get to decide what diff groupings we want to make

```

If we want to create a summary table that contains stats other than counts by group, it's easier to use group_by() +n()

```{r}

# only interested in site, ind of month and year
# we don't just want a count, want summ stats as well

lobster_summary <- lobster_parse_date %>%
  group_by(site) %>%
  summarize(
    mean_size= mean(size_mm, na.rm= TRUE),
    sd_size= sd(size_mm, na.rm= TRUE),
    lobster_number= n()
  )

# another option is tally, but have to give group_by() first, then do not need to write out summarize

```

#### Find confidence intervals

Use t-test() fxn to find conf int for one sample and perform t tests to compare means of two samples

```{r}

# if calculate 95% conf int: means if took more samples from pop that in 95% of intervals reported we would expect to find actual pop mean

# will be using t dist, not z
# t dist is normal dist that acounts for increased uncertainty as sample size gets smaller, with n<30 z and t very sim

# want some meas of uncertainty

ivee_lobsters <- lobster_tidy %>%
  filter(site== "IVEE") %>% 
  pull(size_mm)

# pull() creates vector instead of df, call in console to view-- need vector for t.test()

t.test(ivee_lobsters)

# t score is measure of how diff sample score is from claim, p val tells us if a null hyp is true, the probability that could've taken random sample of pop is mean=0-- very small probability
# most interested in CI: if i took a bunch more samples from pop, I would expect 95% of samples to contain true pop mean

```

#### Two-sample t-test to compare means

We want to ask if there is a sig diff in lobster lengths at Naples and Mohawk reefs?

We've done our nec exploratory analyses to determine that a two-sample t-test for means comparison is approp

```{r}
# 2 ways to do, same output

# 1: create vectors

napl_sample <- lobster_tidy %>% 
  filter(site== "NAPL") %>% 
  pull(size_mm)

mohk_sample <- lobster_tidy %>% 
  filter(site== "MOHK") %>% 
  pull(size_mm)

mn_ttest <- t.test(napl_sample, mohk_sample)

mn_ttest

```

There is a signif diff in lobster lengths btwn Naples and Mohawk Reef -- sig diff (p val) should be least interesting part reported (t(df)= statistic, p< .001, alpha=.05)

```{r}
# 2: create df
# using model format-- less tedious and also get thinking about regression

lobster_mn <- lobster_tidy %>% 
  filter(site %in% c("NAPL", "MOHK"))

mn_ttest2 <- t.test(size_mm ~ site, data= lobster_mn)

mn_ttest2
# same results as separated vector form
# p val tells very unlikely that means would be this diff if they were drawn from same pop

mn_ttest2$p.value
mn_ttest2$statistic
mn_ttest2$conf.int

# want to include these in actual text
```

There is a sig diff (t(1850.8)= -19.849, p= 2.2e-16) -- issues w reproducibility, if anything in the above code were to change, these outputs would change

Here is the p-value: `r mn_ttest2$p.value`

Here is my t-stat: `r mn_ttest2$statistic`

#### Now a heatmap!

```{r}

lobster_ys <- lobster_parse_date %>% 
  count(obs_year, site)

ggplot(data= lobster_ys, aes(x= obs_year, y= site)) +
  geom_tile(aes(fill= n))

```

